---
title: "ProteoMaker Pipeline Parameters"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

find_parameter_yaml <- function() {
  candidates <- c(
    file.path("..", "inst", "config", "parameters.yaml"),
    file.path("inst", "config", "parameters.yaml"),
    system.file("config", "parameters.yaml", package = "ProteoMaker")
  )
  candidates <- candidates[nzchar(candidates)]
  candidates[file.exists(candidates)][1]
}

as_text <- function(x) {
  if (is.null(x) || length(x) == 0 || all(is.na(x))) return("NA")
  if (is.list(x)) x <- unlist(x, use.names = TRUE)
  paste(x, collapse = ", ")
}

fmt_default <- function(x) {
  x <- as_text(x)
  ifelse(nchar(x) > 40, paste0(substr(x, 1, 37), "..."), x)
}

params <- yaml::yaml.load_file(find_parameter_yaml())$params
param_df <- do.call(rbind, lapply(names(params), function(parameter) {
  p <- params[[parameter]]
  data.frame(
    Parameter = parameter,
    Section = p$group,
    List = p$type,
    Class = p$class,
    Default = fmt_default(p$default),
    Range = paste(as_text(p$min), as_text(p$max), sep = " - "),
    Choices = if (is.null(p$choices)) "" else as_text(p$choices),
    Description = p$description,
    stringsAsFactors = FALSE
  )
}))

list_labels <- c(
  paramGroundTruth = "Ground Truth and Experimental Design",
  paramProteoformAb = "Proteoform Abundance",
  paramDigest = "Digestion and Sample Preparation",
  paramMSRun = "MS Run and Search",
  paramDataAnalysis = "Data Analysis"
)

list_purpose <- c(
  paramGroundTruth = "Define the input proteome, experiment layout, expressed proteins, and PTM ground truth.",
  paramProteoformAb = "Generate quantitative proteoform abundances and regulation patterns.",
  paramDigest = "Digest proteoforms and optionally simulate PTM enrichment or sample preparation losses.",
  paramMSRun = "Simulate detection, missing values, instrument noise, false IDs, and localization errors.",
  paramDataAnalysis = "Control protein summarisation and statistical testing."
)

parameter_table <- function(df) {
  out <- df[, c("Parameter", "Default", "Range", "Description"), drop = FALSE]
  knitr::kable(out, align = c("l", "l", "l", "l"), row.names = FALSE)
}
```

```{=html}
<style>
table {
  width: 100%;
  table-layout: fixed;
  border-collapse: collapse;
}
th, td {
  vertical-align: top;
  padding: 0.35rem 0.5rem;
  overflow-wrap: anywhere;
  word-break: break-word;
}
table td:nth-child(1), table th:nth-child(1) { width: 30%; }
table td:nth-child(2), table th:nth-child(2) { width: 8%; }
table td:nth-child(3), table th:nth-child(3) { width: 10%; }
table td:nth-child(4), table th:nth-child(4) { width: 52%; }
code { white-space: normal; overflow-wrap: anywhere; word-break: break-word; }
</style>
```

This vignette describes the parameters used by ProteoMaker. The parameters are split into two groups:

- `set_proteomaker()` controls paths, parallelization, and which downstream work is run.
- `def_param()` returns the simulation parameters used by the pipeline stages.

## Parameter Flow

```{mermaid}
flowchart LR
  config["set_proteomaker()<br/>paths and runtime"]
  params["def_param()<br/>simulation settings"]
  run["run_sims()"]
  gt["Ground truth"]
  sim["Abundance, digest, MS"]
  out["Analysis and benchmarks"]

  config --> run
  params --> run
  run --> gt --> sim --> out
```

## Summary

### Configuration from `set_proteomaker()`

These parameters are not part of the simulation hash. They control where files are read and written, how much parallelism is used, and whether optional downstream steps run.

| Parameter | Default | Purpose |
|------------------------|-----------------------:|------------------------|
| `fastaFilePath` | package `Proteomes` folder | Directory containing FASTA files referenced by `PathToFasta`. |
| `resultFilePath` | `SimulatedDatasets` | Output directory for cached simulation stages and benchmark results. |
| `cores` | `2` | Number of CPU cores used by parallel steps. |
| `clusterType` | `PSOCK` | Parallel cluster backend, usually `PSOCK` or `FORK`. |
| `runStatTests` | `TRUE` | Run statistical tests after simulation. |
| `calcAllBenchmarks` | `TRUE` | Calculate benchmark summaries after statistical testing. |

### Simulation Parameters from `def_param()`

```{r summary-table}
summary_df <- do.call(rbind, lapply(names(list_labels), function(list_name) {
  data.frame(
    List = list_name,
    Section = list_labels[[list_name]],
    Parameters = sum(param_df$List == list_name),
    Purpose = list_purpose[[list_name]],
    stringsAsFactors = FALSE
  )
}))
knitr::kable(summary_df, align = c("l", "l", "r", "l"), row.names = FALSE)
```

### Section Map

```{r section-map}
section_df <- aggregate(
  Parameter ~ List + Section,
  data = param_df,
  FUN = length
)
names(section_df)[names(section_df) == "Parameter"] <- "Parameters"
section_df$List <- factor(section_df$List, levels = names(list_labels))
section_df <- section_df[order(section_df$List, section_df$Section), ]
section_df$List <- as.character(section_df$List)
knitr::kable(section_df, align = c("l", "l", "r"), row.names = FALSE)
```

## Using the Parameters

Use `set_proteomaker()` once for runtime configuration and `def_param()` for the simulation design.

```{r usage-example, echo=TRUE, eval=FALSE}
config <- set_proteomaker(
  fastaFilePath = system.file("Proteomes", package = "ProteoMaker"),
  resultFilePath = file.path(tempdir(), "SimulatedDataSets"),
  cores = 1,
  clusterType = "PSOCK",
  runStatTests = TRUE,
  calcAllBenchmarks = TRUE
)

Param <- def_param()
Param$paramGroundTruth$PathToFasta <- "fasta_example.fasta"
Param$paramGroundTruth$NumCond <- 2
Param$paramGroundTruth$NumReps <- 3
```

Each top-level element in `Param` corresponds to one pipeline stage. Parameters can be vectors; `run_sims()` expands them into combinations.

## `set_proteomaker()` Configuration

### Paths

| Parameter | Description |
|------------------------------------|------------------------------------|
| `fastaFilePath` | Directory containing FASTA files. `PathToFasta` values from `Param$paramGroundTruth` are resolved relative to this directory. |
| `resultFilePath` | Directory used for saved stage files, retrieved simulations, and benchmark tables. The directory is created if needed. |

### Runtime

| Parameter | Description |
|------------------------------------|------------------------------------|
| `cores` | Number of cores used by parallelized steps. Larger values can substantially increase RAM use. |
| `clusterType` | Parallel backend. `PSOCK` is portable; `FORK` can be faster on Unix-like systems but is not available everywhere. |
| `runStatTests` | If `FALSE`, simulation can stop before statistical testing and occupancy/benchmark calculation. |
| `calcAllBenchmarks` | If `TRUE`, benchmark metrics are calculated after statistical testing. |

## `def_param()` Simulation Parameters

The tables below follow the actual top-level lists returned by `def_param()`. Defaults and ranges are read from `inst/config/parameters.yaml`.

```{r parameter-sections, results="asis"}
for (list_name in names(list_labels)) {
  cat("\n### `", list_name, "`: ", list_labels[[list_name]], "\n\n", sep = "")
  cat(list_purpose[[list_name]], "\n\n")

  list_df <- param_df[param_df$List == list_name, , drop = FALSE]
  for (section in unique(list_df$Section)) {
    cat("\n#### ", section, "\n\n", sep = "")
    print(parameter_table(list_df[list_df$Section == section, , drop = FALSE]))
    cat("\n")
  }
}
```

## PTM Setup Recipe

This minimal phosphorylation setup keeps the unmodified counterpart and creates approximately one modified site per modified proteoform.

```{r ptm-recipe, echo=TRUE, eval=FALSE}
Param$paramGroundTruth$FracModProt <- 0.5
Param$paramGroundTruth$PTMTypes <- list(mods = c("ph"))
Param$paramGroundTruth$PTMTypesMass <- list(ph = 79.966331)
Param$paramGroundTruth$PTMTypesDistr <- list(ph = 1)
Param$paramGroundTruth$ModifiableResidues <- list(mods = list(ph = c("S", "T", "Y")))
Param$paramGroundTruth$ModifiableResiduesDistr <- list(
  mods = list(ph = c(S = 0.86, T = 0.13, Y = 0.01))
)
Param$paramGroundTruth$PTMMultipleLambda <- 0
Param$paramGroundTruth$PropModPerProt <- 1
Param$paramGroundTruth$RemoveNonModFormFrac <- 0
```

## Practical Notes

- Use `set_proteomaker()` for filesystem and runtime settings.
- Use `def_param()` for anything that should define a simulated dataset.
- Keep small parameter grids while testing; each vector-valued parameter multiplies the number of simulations.
- `PathToFasta` names a FASTA file, while `fastaFilePath` names the directory that contains it.
