---
title: "PTM Occupancy Ground Truth"
author: ""
date: "2026-05-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PTM Occupancy Ground Truth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette runs a small ProteoMaker simulation with phosphorylation-like PTMs and compares the PTM occupancies estimated from peptide-level data with the ground-truth occupancies calculated from proteoform abundances.

The estimated occupancy table reports peptide-local PTM positions and protein-level PTM positions. The comparison below joins estimated and ground-truth occupancies by accession and protein-level PTM site.

```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Configure a Small PTM Simulation

```{r}
library(ProteoMaker)

proteomaker_config <- set_proteomaker(
  fastaFilePath = system.file("Proteomes", package = "ProteoMaker"),
  resultFilePath = file.path(tempdir(), "ProteoMaker_PTM_occupancy"),
  cores = 1,
  clusterType = "PSOCK",
  runStatTests = TRUE,
  calcAllBenchmarks = FALSE
)

Param <- def_param()

Param$paramGroundTruth$PathToFasta <- "fasta_example.fasta"
Param$paramGroundTruth$NumCond <- 2
Param$paramGroundTruth$NumReps <- 3
Param$paramGroundTruth$PercExpressedProt <- 1

# Generate modified proteins and keep their unmodified counterparts.
Param$paramGroundTruth$FracModProt <- 0.5
Param$paramGroundTruth$PropModPerProt <- 1
Param$paramGroundTruth$PTMMultipleLambda <- 0.0
Param$paramGroundTruth$RemoveNonModFormFrac <- 0

Param$paramGroundTruth$PTMTypes <- list(mods = c("ph"))
Param$paramGroundTruth$PTMTypesMass <- list(ph = 79.966331)
Param$paramGroundTruth$PTMTypesDistr <- list(ph = 1)
Param$paramGroundTruth$ModifiableResidues <- list(mods = list(ph = c("S", "T", "Y")))
Param$paramGroundTruth$ModifiableResiduesDistr <- list(
  mods = list(ph = c(S = 0.86, T = 0.13, Y = 0.01))
)

# Keep the MSRun simple so the comparison is easy to inspect.
Param$paramMSRun$PercDetectability <- 1
Param$paramMSRun$PercDetectedVal <- 1
Param$paramMSRun$MSNoise <- 0
Param$paramMSRun$WrongIDs <- 0.0
Param$paramMSRun$WrongLocalizations <- 0.0
Param$paramMSRun$MaxNAPerPep <- 100
```

## Run the Simulation and Retrieve Outputs

```{r}
allBs <- run_sims(Param, proteomaker_config, overwrite = TRUE)

sim_param <- allBs[[1]]$Param
data_analysis <- get_simulation(sim_param, proteomaker_config, stage = "DataAnalysis")
proteoform_ab <- get_simulation(sim_param, proteomaker_config, stage = "ProteoformAb")$proteoformAb

estimated_occ <- data_analysis$Occupancies
ground_truth_occ <- calcGroundTruthPTMOccupancy(proteoform_ab, sim_param)
```

## Site Counts

```{r}
condition_cols <- paste0("C_", seq_len(sim_param$NumCond))
condition_replicates <- split(sim_param$QuantColnames, rep(condition_cols, each = sim_param$NumReps))

for (cond in condition_cols) {
  ground_truth_occ[[cond]] <- rowMeans(
    ground_truth_occ[, condition_replicates[[cond]], drop = FALSE],
    na.rm = TRUE
  )
}

estimated_sites <- data.frame(
  Accession = vapply(estimated_occ$Accession, function(x) unlist(x)[1], character(1)),
  PTMType = vapply(estimated_occ$PTMType, function(x) unlist(x)[1], character(1)),
  PTMPos = vapply(estimated_occ$ProteinPTMPos, function(x) as.integer(unlist(x)[1]), integer(1)),
  estimated_occ[, condition_cols, drop = FALSE]
)

ground_truth_sites <- data.frame(
  Accession = vapply(ground_truth_occ$Accession, function(x) unlist(x)[1], character(1)),
  PTMType = vapply(ground_truth_occ$PTMType, function(x) unlist(x)[1], character(1)),
  PTMPos = vapply(ground_truth_occ$PTMPos, function(x) unlist(x)[1], integer(1)),
  ground_truth_occ[, condition_cols, drop = FALSE]
)

comparison <- merge(
  estimated_sites,
  ground_truth_sites,
  by = c("Accession", "PTMType", "PTMPos"),
  suffixes = c("_estimated", "_ground_truth")
)

site_cols <- c("Accession", "PTMType", "PTMPos")
estimated_site_ids <- unique(estimated_sites[, site_cols])
ground_truth_site_ids <- unique(ground_truth_sites[, site_cols])
matched_site_ids <- unique(comparison[, site_cols])

counts <- data.frame(
  Metric = c(
    "Ground-truth PTM sites",
    "Estimated PTM sites",
    "Matched PTM sites",
    "Ground truth only",
    "Estimated only"
  ),
  Count = c(
    nrow(ground_truth_site_ids),
    nrow(estimated_site_ids),
    nrow(matched_site_ids),
    nrow(ground_truth_site_ids) - nrow(matched_site_ids),
    nrow(estimated_site_ids) - nrow(matched_site_ids)
  )
)

counts
```

## Occupancy Distributions

The check is whether estimated and ground-truth occupancies occupy similar ranges. Ground-truth distributions are shown after averaging replicates within each condition.

```{r}
estimated_distribution <- summary(estimated_occ[, condition_cols, drop = FALSE])
ground_truth_distribution <- summary(ground_truth_occ[, condition_cols, drop = FALSE])

estimated_distribution
ground_truth_distribution
```

```{r fig.width=7, fig.height=4}
# histograms 
for(i in condition_cols) {
  hist(estimated_occ[[i]], main = paste("Estimated occupancy -", i), xlab = "Occupancy", 
       breaks = 20*diff(range(estimated_occ[[i]])), xlim = c(-1, 2), border = NA, col = "#1b9e77")
  hist(ground_truth_occ[[i]], main = paste("Ground truth occupancy -", i), xlab = "Occupancy", 
       xlim=c(-1,2), breaks = 20, border = NA, col="#d95f02")
}
```

## Direct Value Comparison

For matched PTM sites, compare the estimated and true occupancy values per condition. 

```{r}
for (i in condition_cols) {
  plot(
    comparison[[paste0(i, "_estimated")]],
    comparison[[paste0(i, "_ground_truth")]],
    main = paste("Estimated vs Ground Truth Occupancy -", i),
    xlab = "Estimated Occupancy",
    ylab = "Ground Truth Occupancy",
    pch = 16, col = rgb(0.1, 0.1, 0.1, 0.5)
  )
  abline(a = 0, b = 1, col = "red", lty = 2)
}

# on log-scale
for (i in condition_cols) {
  plot(
    log10(comparison[[paste0(i, "_estimated")]] + 1e-6),
    log10(comparison[[paste0(i, "_ground_truth")]] + 1e-6),
    main = paste("Estimated vs Ground Truth Occupancy (log10) -", i),
    xlab = "Estimated Occupancy (log10)",
    ylab = "Ground Truth Occupancy (log10)",
    pch = 16, col = rgb(0.1, 0.1, 0.1, 0.5)
  )
  abline(a = 0, b = 1, col = "red", lty = 2)
}

```

