---
title: "PhosFake Vignette"
author: ""
date: "2024-07-25"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PhosFake Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

PhosFake is a simulation tool for generating synthetic proteomics datasets. This vignette will guide you through the process of setting up and running PhosFake simulations, as well as visualizing the results.

## Installation

PhosFake requires installation via the `devtools` package. You can install PhosFake using the following command:

```{r}
# Install PhosFake
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools", repos = "https://cran.r-project.org")
}
devtools::install_github("veitveit/PhosFake")
```

And load the library
```{r}
# Load PhosFake
library(PhosFake)
```

## Configuration

Set up the paths and directories for input and output data, and configure the computational settings. 
This step ensures that the necessary directories exist and the computational resources are configured correctly. The fasta file path is set to the Proteomes folder in the PhosFake package with the following
proteomes: _fasta_example.fasta, fasta_full_human.fasta, fasta_full_mouse.fasta, fasta_full_yeast.fasta_

In this case, the results will be written in the temporary folder. In case you want to run many simulations, we recommend keeping the results file in a different folder by changing the argument `resultFilePath`.


```{r}
# Configure the paths and settings
phosfake_config <- set_phosfake(
  # Default path to the Proteomes folder in the PhosFake package
  fastaFilePath =  system.file("Proteomes", package = "PhosFake"),
  # Default path to the temporary directory
  resultFilePath = paste0(tempdir(), "/SimulatedDataSets"),
  # Increase to the number of cores you want to use. 
  # Large numbers can lead to considerably higher RAM usage
  cores = 1, 
  # Should work on all operation systems. If you have problems, try "FORK" instead
  clusterType = "PSOCK",
  # Run statistical tests on the simulated data to compare ground truth with simulated values
  runStatTests = TRUE,
  # Calculate all benchmarks for the simulated data
  calcAllBenchmarks = TRUE
)
```

## Set Simulation Parameters

Generate a default list of testing parameters using the `def_param` function. You can overwrite the default values with those you want to test by changing the parameters directly, or by loading a yaml file. An example 
for the yaml file with the default parameters is available [here](https://github.com/veitveit/PhosFake/blob/master/inst/config/parameters.yaml). 

A more details documentation of the parameters can be found [here](https://computproteomics.bmb.sdu.dk/app_direct/PhosFake/Parameters.html).

Multiple values for each parameter can be combined to generate different simulated datasets. This flexibility allows you to customize the simulations according to your arrays of experimental needs.

```{r}
# Generate default parameters
Param <- def_param()


# Example of overwriting default values
Param$paramGroundTruth$PathToFasta <- "fasta_example.fasta"
Param$paramGroundTruth$PercExpressedProt <- 1.0
# Param$paramGroundTruth$NumReps <- c(3)
# Param$paramGroundTruth$NumCond <- 5
# Param$paramGroundTruth$FracModProt <- 0.5
# Param$paramGroundTruth$PTMTypes <- "ph"
# Param$paramGroundTruth$PTMTypesMass <- c(79.966331)
# Param$paramGroundTruth$PTMTypesDist <- c(1)
# Param$paramGroundTruth$PTMMultipleLambda <- c(0.1)
# Param$paramGroundTruth$ModifiableResidues <- list(c("S", "T", "Y"))
# Param$paramGroundTruth$ModifiableResiduesDistr <- list(c(0.86,0.13, 0.01))
Param$paramGroundTruth$NumReps <- c(3:5)
# Param$paramProteoformAb$QuantNoise <- seq(0.1, 0.9, 0.5)
# Param$paramProteoformAb$DiffRegFrac <- c(0.1, 0.3, 0.5)
# Param$paramProteoformAb$DiffRegMax <- seq(0.5, 2, 0.5)
# Param$paramDigest$Enzyme <- "trypsin"
# Param$paramDigest$PropMissedCleavages <- 0.01
# Param$paramDigest$MaxNumMissedCleavages <- 4
# Param$paramDigest$PepMinLength <- 7
# Param$paramDigest$PepMaxLength <- 30
# Param$paramMSRun$PercDetectedPep <- seq(0.1, 0.5, 0.1)
# Param$paramDataAnalysis$MinUniquePep <- 100

```

## Running Simulations

Run the simulations using the defined parameters and configuration. This step generates the synthetic datasets based on the specified parameters.

```{r}
# Run the simulations
allBs <- run_sims(Param, phosfake_config)

```

## Retrieving Results

Get the results of an individual simulation. This allows you to inspect the parameters and outputs of a specific run. Retrieving results is useful for detailed analysis of a particular simulation.

```{r}
# Retrieve results}
res <- get_simulation(allBs[[1]]$Param, phosfake_config)

```

## Retrieving Benchmarks

Create a matrix of benchmarks from all simulations and save it as a CSV file. This file contains various metrics and results from the simulations. Generating benchmarks helps in comparing different simulations and assessing their performance.

```{r}
# Generate the benchmark matrix
benchmarks <- matrix_benchmarks(allBs, phosfake_config)
write.csv(benchmarks, file = paste0(phosfake_config$resultFilePath, "/allBenchmarks.csv"))

```

## Visualizing Results

Visualize the results of the benchmarks to analyze the performance and outcomes of the simulations. Visualization aids in interpreting the data and identifyinsg trends or anomalies

```{r}
# visualize the benchmarks and parameters of the second simulation
visualize_benchmarks(benchmarks, 2)

```
